---
title: 'Workflow illustration for “Comparing model-based unconstrained ordination methods in the analysis of high-dimensional compositional count data”'
author: "Wenqi Tang"
date: "2025-03-18"
output: 
  github_document:
    pandoc_args: --webtex
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, size="footnotesize", fig.width=6, fig.height=5, 
fig.align="center",dev="png", code.frame = TRUE, warning = FALSE, fig.pos='H', 
message=FALSE, clean=TRUE)
```

This document reproduces the main content of the introduction and simulation section of the paper with the same name by Wenqi Tang, Pekka Korhonen, Jenni Niku,  Klaus Nordhausen and Sara Taskinen. It includes microbiome data visualization and example about how to handle compositional data and apply ordination methods in R within the simulation section. We primarily introduce the three distinct ordination methods mentioned in the paper, which are applicable to high-dimensional compositional count data. Specifically, we explore classical ordination methods, those based on generalized linear latent variable models (GLLVMs), and copula-based methods, applying them to microbial data as detailed in @Kumar2017. The dataset used in this analysis consists of a 56 × 985 matrix, representing 985 bacterial species sampled from 56 soil sites across three distinct climatic regions: Ny-Ålesund (high Arctic), Kilpisjärvi (low Arctic), and Mayrhofen (European Alps) [@Kumar2017]. Due to biological or technical factors, such datasets often contain a significant proportion of zero observations (e.g., 63% zeros in the microbiome data). Additionally, microbiome data exhibit a compositional nature, as highlighted in @Glooretal2016, meaning the data represent relative abundances rather than absolute counts.

## Data preprocessing

We order the columns according to sparseness and use a heatmap to visualize the sparsity of the data. Additionally, a mean-variance plot is employed to assess the overdispersed in the data.

```{r data, echo=TRUE}
library(robCompositions)
library(vegan)
library(devtools) # Load devtools for package installation
#devtools::install_github("JenniNiku/gllvm") # Install the latest gllvm package from GitHub
library(gllvm) # Load the gllvm package
source_url("https://raw.githubusercontent.com/tangwenq/microbial-data/main/GCLVM.R")
library(reshape2) # For data reshaping
library(scales)

# Base the simulations to arctic microbialdata from Nissinen et al.
# Data included in gllvm package

data("microbialdata")
data <- microbialdata$Y
# order columns according to sparseness
data.s <- cbind(1:dim(data)[2], apply(data == 0, 2, sum))
ind <- data.s[order(data.s[, 2], decreasing = FALSE), 1]
data <- data[, ind]
dim(data)

# Draw mean-variance plot, (shown in right picture of Figure 1)

# Compute mean and variance for each column
means <- apply(data, 2, mean) # Column-wise means
variances <- apply(data, 2, var) # Column-wise variances

# Define a Negative Binomial (NB) variance function
nb_model <- function(mean, theta) {
  mean + mean^2 * theta # NB variance formula
}

# Fit the NB model to the data using nonlinear least squares
fit <- nls(variances ~ nb_model(means, theta),
  start = list(theta = 0.1)
) # Use an initial guess for theta

# Extract the optimal dispersion parameter (theta)
theta_opt <- coef(fit)["theta"] # Extract the estimated theta value
cat("Optimal dispersion parameter (theta):", theta_opt, "\n")


# Plot mean-variance relationship
plot(means, variances,
  xlab = "Mean", # X-axis label
  ylab = "Variance", # Y-axis label
  pch = 16,
  cex = 0.7, # Point size
  col = "black",
  xlim = c(0, 30), # Limit for X-axis
  ylim = c(0, 1000), # Limit for Y-axis
  cex.lab = 1.5, # Label font size
  cex.axis = 1.4, # Axis font size
  cex.main = 1.6
) # Title font size

# Add a Poisson reference line (variance = mean)
abline(a = 0, b = 1, col = "blue", lwd = 2) # Poisson variance line

# Add the Negative Binomial relationship line
theta <- theta_opt # Use the estimated dispersion parameter
nb_variance <- means + means^2 * theta # Compute NB variance
lines(sort(means), sort(nb_variance),
  col = "blue", lwd = 2, lty = 2
) # Add the NB variance line (dashed)


# Draw heatmap of data, (shown in left picture of Figure 1)

# Convert data to long format for plotting
long_data <- melt(data, varnames = c("Site", "Species"), value.name = "Richness")


# Generate the heatmap
ggplot(long_data, aes(x = Species, y = Site, fill = Richness)) +
  geom_tile() +
  scale_fill_gradientn(
    colors = rev(hcl.colors(n = 8, palette = "Spectral")), # Reverse gradient colors
    values = rescale(c(0, 2, 5, 10, 15, 50, 100, 200, 1024)), # Emphasize 0-15 range
    oob = scales::squish, # Handle out-of-bounds values
    limits = c(0, 1024) # Set gradient range
  ) +
  theme_minimal() + # Minimal theme
  labs(
    x = "Species",
    y = "Sample",
    fill = "Count" # Axis and legend labels
  ) +
  theme(
    axis.text.x = element_text(size = 14, face = "bold", angle = 90, hjust = 1), # Bold x-axis text
    axis.text.y = element_blank(), # Remove y-axis labels
    axis.ticks = element_blank(), # Remove axis ticks
    axis.title.x = element_text(size = 18),
    axis.title.y = element_text(size = 18),
    legend.title = element_text(size = 15),
    legend.text = element_text(size = 9),
    legend.position = "bottom",
    legend.direction = "horizontal"
  ) +
  scale_x_discrete(
    breaks = c(50, 100, 200, 400), # Specify discrete breaks if Species are numbered
    labels = c("50", "100", "200", "400") # Custom labels
  ) +
  geom_vline(
    xintercept = c(50, 100, 200, 400),
    color = "black",
    linetype = "solid", # Use solid lines for emphasis m = 50, 100, 200, 400
    size = 1
  ) +
  guides(
    fill = guide_colorbar(
      barwidth = 32,
      barheight = 2,
      breaks = seq(0, 250, by = 50)
    )
  )

```

## Classical ordination

Classical methods for unconstrained ordination, such as **non-metric multidimensional scaling (nMDS)** and **principal component analysis (PCA)**, are widely used due to their computational efficiency and ease of implementation. Compositional data are subject to sum constraints, require **centered log-ratio (clr) transformation** in **R package `robCompositions`** to map compositional data from the Aitchison space to real space before applying these methods. For **nMDS**, the Euclidean distance can be used to evaluate the clr-transformed data by specifying `distance = "euclidean"`. Alternatively, for the original compositional data, dissimilarity measures such as the Bray-Curtis distance `distance = "bray"` can be applied using the **R package `vegan`**. As notable downsides, classical methods lack a probabilistic framework, and may have challenges in handling zero values.

### Example

In the following analysis, we demonstrate the application of classical ordination methods to compositional data.

```{r, classical, echo=TRUE}
y <- data
# Distance-based methods: nMDS

    fit_mds <- metaMDS(y, distance = "bray", trace = FALSE)
    mds_ords <- scale(fit_mds$ points) # extract ordinations from nMDS
    
    # PCA and nMDS on CLR-transformed data
    clr_y <- cenLR(y + 1) # CLR transformation of sim_y with added 1

    fit_clrmds <- metaMDS(clr_y$x.clr, distance = "euclidean", autotransform = FALSE, noshare = FALSE, wascores = FALSE, trace = FALSE)
    clrmds_ords <- scale(fit_clrmds$points)
    
    fit_pca <- prcomp(clr_y$x.clr, scale = TRUE) # PCA on the CLR-transformed data clr_y
    pca_ords <- scale(fit_pca$x[, 1:2]) # extract ordinations from PCA
```

The microbiome data was collected from three regions: "Mayrhofen," "Kilpisjarvi," and "Ny-Ålesund." To analyze the community structure of these data, we can visualize the ordinations extracted using classical methods by plotting ordination plots. The following code demonstrates how to perform ordination analysis on CLR-transformed data using PCA as an example and visualize the results.

```{r classical_vis, echo=TRUE,message=FALSE}


# Extract unique regions and symbols
regions <- unique(microbialdata$X$Region) # Unique region names
pch_values <- as.numeric(factor(regions)) + 15 # Define symbols manually: circle (15), triangle (16), plus (17)
colors <- c("black", "blue", "grey")[as.numeric(factor(regions))] # Set specific colors for each region

region_names <- c("Mayrhofen", "Kilpisjarvi", "Ny-Alesund")

# Draw the ordination plot for pca
plot(pca_ords[, 1], pca_ords[, 2], type = "n", 
     xlab = "PC1", ylab = "PC2", main = "PCA of CLR-transformed Data")
points(pca_ords[, 1], pca_ords[, 2], col = colors[as.numeric(factor(regions))], 
       pch = pch_values[as.numeric(factor(regions))], cex = 1.2) 

legend("topleft",                      
       legend = region_names,       
       pch = pch_values,               
       col = colors,                    
       ncol = 3,                        
       cex = 1)                       
```

## Ordination based on latent variable models (GLLVM)

Generalized linear latent variable models (GLLVMs) provide a flexible framework for jointly modeling multivariate response data, such as microbial count data. GLLVMs extend generalized linear models (GLMs) by incorporating correlations among response variables using a factor-analytic approach. This allows GLLVMs to be used for model-based ordination for any response type. Assume $\boldsymbol{u}_i = (u_{i1}, \dots, u_{id})^\top$ is a $d$-dimensional latent variable following a standard multivariate normal distribution. In GLLVMs, it is assumed that, conditional on the latent variables $\boldsymbol{u}_i$, the response variables $y_{ij}$ are independently distributed according to some distribution $F(\mu_{ij}, \boldsymbol{\phi})$, where $\mu_{ij} = \mathbb{E}(y_{ij} | \boldsymbol{u}_i)$ is the conditional mean, and $\boldsymbol{\phi}$ includes possible response-specific parameters (e.g., dispersion or zero-inflation parameters). When GLLVMs are used for model-based ordination, $\mu_{ij}$ is linked to the linear predictor via: 
$$ g(\mu_{ij}) = \alpha_i + \beta_{0j} + \boldsymbol{\lambda}_j^\top \boldsymbol{u}_i,$$ where: 

* $g(\cdot)$ is a known link function (typically the log-link for count data). 
* $\beta_{0j}$ is a column-specific intercept to account for differences in column totals.
* $\boldsymbol{\lambda}_j = (\lambda_{j1}, \dots, \lambda_{jd})^\top$ is a$d$ -dimensional vector of factor loadings. 
* $\alpha_i$ is a row-specific intercept to account for differences in row totals.

To ensure model identifiability, the upper triangular part of the factor loading matrix $\boldsymbol{\Lambda} = [\boldsymbol{\lambda}_1 \cdots \boldsymbol{\lambda}_m]^\top$ is set to zero, and its diagonal elements are set to positive values. The `gllvm` package implements the **zero-inflated negative binomial (ZINB)** (`family = "ZINB"`) and **negative binomial (NB)** (`family = "negative.binomial"`) GLLVMs to better handle datasets with a large number of zeros. These models are particularly suited for overdispersed count data, where excess zeros are common. For non-normal response data, estimation is typically based on pproximate marginal likelihood using variational approximations (VA) or Laplace's approximation (LA). 

### Example

To speed up computationsm, one could first, enable parallelization for `gllvm` e.g., with:
```{r parallel, results=FALSE, message=FALSE, eval=FALSE}
TMB::openmp(n = parallel::detectCores()-1, autopar = TRUE, DLL = "gllvm")
```

Then, to fit the model:
```{r GLLVM, echo=TRUE}
# Extract and scale environmental covariates
X <- scale(microbialdata$Xenv[, c("SOM", "pH", "Phosp")])

# Uncomment to fit unconstrained model
#ungllvm_NB <- gllvm(
#  y = y, family = "negative.binomial", sd.errors = FALSE,
#  row.eff = "fixed", num.lv = 2, seed = 123
#)

# Concurrent ordination model
gllvm_NB <- gllvm(
  y = y, X=X,family = "negative.binomial", sd.errors = FALSE,
  row.eff = "fixed", num.lv.c = 2, seed = 123
)

head(scale(gllvm_NB$lvs)) # extract ordination scores
```

Through the `summary()`, `AIC()` or `BIC()` functions, we can observe the AIC and BIC values of the model, which can also be found in **Table 2** of the paper.

```{r gllvm_analysis, echo=TRUE}
AIC(gllvm_NB)
BIC(gllvm_NB)
```

Additionally,

* The estimate pH(CLV1) = -0.135067, p < 0.001, indicates that pH has a **negative influence** on the first latent variable (CLV1), and the p-value (\< 0.001) suggests that this effect is **statistically significant**.
* The estimate pH(CLV2) = 0.034117, p < 0.001, indicates that pH has a **positive influence** on the second latent variable (CLV2), and the p-value (\< 0.001) suggests that this effect is **statistically significant**.

From these results, we can conclude that the **NB-GLLVM model successfully captures the effects of environmental variables (particularly pH) on microbial community structure**. This conclusion is also supported by the following ordination plot, where pH is represented by dark red arrows, indicating its strong influence.

Furthermore, it is worth noting that some environmental variables (e.g., SOM) do not show significant effects on the latent variables. The row effects capture the **site-specific variability** among sampling sites. The p-values reflect whether the influence of each sampling site on the response variable is significant.

For model-based methods, we can use the following code in `gllvm` to generate residuals plots and ordination plots for analysis, assessing the model's fitting performance.

```{r GLLVM_vis, echo=TRUE}


# Visualize residuals for  NB-GLLVM model(reproduces Figure 8)
# the D-S of concurrent NB-GLLVM model, shown in bottom row in Figure 8
par(mar = c(6, 6, 0, 0) + 0.1) # Set plot margins
plot(gllvm_NB,
  which = 1, #when which = 1, it produces a residuals plot
  caption = " ",
  var.colors = 1,
  cex.lab = 2.5,
  cex.axis = 1.5,
  cex.main = 2.5,
  cex = 1
)
par(mar = c(6, 6, 0, 0) + 0.1) # Set plot margins
# Q-Q plot of residuals
plot(gllvm_NB,
  which = 2, #when which = 2, it generates a QQ norm plot
  caption = " ", 
  var.colors = 1, 
  cex.lab = 2.5, 
  cex.axis = 1.5, 
  cex.main = 2.5, 
  cex = 1
) 

# Ordination plots models (reproduces Figure 5)
ordiplot(gllvm_NB,
  which.lvs = 1:2,
  s.colors = c("black", "blue", "grey")[as.numeric(factor(microbialdata$X$Region))],
  rotate = TRUE,
  symbols = TRUE,
  pch = as.numeric(factor(microbialdata$X$Region)) + 15,
  main = "",
  ann = FALSE
) # Set symbols based on region

title(xlab = "ordination score 1", ylab = "ordination score 2", cex.lab = 1.3)
# Add a legend with specified colors and symbols
legend("topleft", # Legend position
  legend = region_names, # Legend labels
  pch = pch_values, # Corresponding symbols
  col = colors, # Corresponding colors (black, blue, grey)
  ncol = 3,
  cex = 1
) # Arrange legend in 3 columns


```

In ordination plot, longer arrows represent covariates with the largest relative effects, and dark red arrows (here associated to pH) indicate covariates with a significant effect to ordination.

## Ordination based on Gaussian copulas (GCLVM)

Copula models couple marginal models for the data with a multivariate model that accounts for covariance across responses. Assume $y_{ij} \sim F_j(\mu_{ij}, \boldsymbol{\phi})$, where $\mu_{ij} = \mathbb{E}(y_{ij})$, and $\boldsymbol{\phi}$ includes possible response-specific parameters. The mean $\mu_{ij}$ is linked to the linear predictor via generalized linear models (GLMs): $$ g(\mu_{ij}) = \alpha_i + \beta_{0j} $$ where $g(\cdot)$ is a known link function, $\alpha_i$ and $\beta_{0j}$ are row-specific and column-specific intercepts,respectively. In the Gaussian copula model, count data $y_{ij}$ are mapped to copula values $z_{ij}$ that follow a multivariate normal distribution: $$F_j(y_{ij} - 1) \leq \Phi(z_{ij}) < F_j(y_{ij})$$ where $F_j(\cdot)$ is the cumulative distribution function (cdf) assumed for the $j$-th column in the data matrix under the marginal GLM, and $\Phi(\cdot)$ is the cdf of the standard normal distribution. For ordination analysis, the copula values $z_{ij}$ are assumed to follow a factor-analytic model: $$ z_{ij} = \boldsymbol{\lambda}_j^\top \boldsymbol{u}_i + \epsilon_{ij}, $$ where: $\boldsymbol{u}_i$ is a $d$-dimensional latent variable associated with the study unit. $\boldsymbol{\lambda}_j$ is a $d$-dimensional vector of factor loadings. $\epsilon_{ij}$ are independent Gaussian errors with variances $\sigma_j^2$. The GCLVM data is generated based on modified code from @Popovic2021. 

The parameters of the copula model are estimated using a two-step procedure:

1. Estimate the marginal distributions $F_j(\cdot)$ using GLMs suitable for sparse, overdispersed count data.
2. Use Monte Carlo expectation-maximization (MCEM) to estimate the covariance parameters in the copula model.

In place of marginal GLMs, due to the compositionality constraint, for  the base model, we use a multivariate GLM with row effects shared by the different species in the data, estimated with the `gllvm` package using the options `num.lv=0` and `row.eff="fixed`. Alternatively, one could include the row effects by treating the data in the long format instead.

### Example

To fit the base model and the copula, use:
```{r GCLVM, echo=TRUE}

# Uncomment for copula ordination based on ZINB distribution
#c_ZINB <- fit_copula(y, gllvm.fam = "ZINB", reff = "fixed", sd.errors = FALSE, seed = 123, lv.n = 0)

# Copula ordination with NB distribution
c_NB <- fit_copula(y[,1:200], gllvm.fam = "negative.binomial", reff = "fixed", sd.errors = FALSE, seed = 123, lv.n = 0)
c_NB_ords <- scale(c_NB$scores) # extract ordinations
head(c_NB_ords)
```

Afterwards, the (unconstrained) ordination plot can be constructed with:
```{r GCLVM_ana, echo=TRUE}
plot(c_NB_ords[, 1], c_NB_ords[, 2], main = "NB_GCLVM ordination plot ",
     xlab = "PC1", ylab = "PC2", cex = 1.2,
     col = colors[as.numeric(factor(regions))], 
     pch = pch_values[as.numeric(factor(regions))])

legend("topleft",                      
       legend = region_names,       
       pch = pch_values,               
       col = colors,                    
       ncol = 3,                        
       cex = 1) 
```

## References

---
references:
- id: Glooretal2016
  title: It's all relative — analyzing microbiome data as compositions
  author:
  - family: Gloor
    given: Gregory B.
  - family: Wu
    given: Jia Rong
  - family: Pawlowsky-Glahn
    given: Vera
  - family: Egozcue
    given: Juan José
  publisher: Annals of Epidemiology
  volume: 26
  number: 5
  page: 322-329
  type: article-journal
  issued:
    year: 2016
  doi: 10.1016/j.annepidem.2016.03.003

- id: Kumar2017
  title: Plants Assemble Species-Specific Bacterial Communities from Common Core Taxa in Three Arcto-Alpine Climate Zones
  author:
  - family: Kumar
    given: Manoj
  - family: Brader
    given: Gerald
  - family: Sessitsch
    given: Angela
  - family: Mäki
    given: Minna
  - family: van Elsas
    given: Jan Dirk
  - family: Nissinen
    given: Riikka
  publisher: Frontiers in Microbiology
  volume: 8
  page: 12
  type: article-journal
  issued:
    year: 2017
  doi: 10.3389/fmicb.2017.00012

- id: Popovic2021
  title: Fast model-based ordination with copulas—Simulation code (v1.0.0)
  author:
  - family: Popovic
    given: Gordana
  type: manual
  issued:
    year: 2021
  doi: 10.5281/zenodo.5525716
---
